---
title: Working with data in a data frame
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=4, fig.path = "fig/dataframes-")
```

# Working with data in a data frame

As we saw earlier, `read.csv` loads tabular data from a CSV file into a data frame.

```{r}
diabetes <- read.csv("data/diabetes.csv")

class(diabetes)
head(diabetes)
colnames(diabetes)
ncol(diabetes)
nrow(diabetes)
```

> ### Tip {.callout}
>
> A data frame can also be created de novo from some column vectors with `data.frame`. For example
>
> ```{r}
> data.frame(foo=c(10,20,30), bar=c("a","b","c"))
> ```
>

> ### Tip {.callout}
>
> A data frame can have both column names (`colnames`) and rownames (`rownames`). However, the modern convention is for a data frame to use column names but not row names. Typically a data frame contains a collection of items (rows), each having various properties (columns). If an item has an identifier such as a unique name, this would be given as just another column.
>

## Indexing data frames

As with a matrix, a data frame can be accessed by row and column with `[,]`.

Internally, a data frame is a list of column vectors. We can use the `$` syntax we saw with lists to access columns by name.


The data we are working with is derived from a dataset called `diabetes` in the `faraway` package. The rows are people interviewed as part of a study of diabetes prevalence. The column `glyhb` is a measurement of Glycosylated Haemoglobin. Values greater than 7 are usually taken as a positive diagnosis of diabetes. Lets add this as a column.

```{r}
diabetes$diabetic <- diabetes$glyhb > 7.0

head(diabetes)
```


## Factors

When R loads a CSV file, it tries to give appropriate types to the columns.
Lets examine what types R has given our data.

```{r}
str(diabetes)
```

We might have expected the text columns to be the "character" data type, but they are instead "factor"s.

```{r}
head( diabetes$frame )
```

R uses factor data type to store a vector of *categorical* data. The different possible categories are called *levels*.

Factors can be created from character vectors with `factor`. We sometimes care what order the levels are in, since this can affect how data is plotted or tabulated by various functions. If there is some sort of baseline level, such as "wildtype strain" or "no treatment", it is usually given first. `factor` has a parameter `levels=` to specify the desired order of levels.

Factors can be converted back to a character vector with `as.character`.

When R loaded our data, it chose levels in alphabetical order. Lets adjust that for the column `diabetes$frame`.

```{r}
diabetes$frame <- factor(diabetes$frame, levels=c("small","medium","large"))

head( diabetes$frame )
```

### Plotting factors

If we `plot` factors, we can see that R uses the order of levels in the plot.

```{r}
plot( diabetes$frame )
plot( diabetes$gender, diabetes$frame )
plot( factor(diabetes$diabetic) )
plot( factor(diabetes$diabetic), diabetes$frame )
```

### Summarizing factors

The `table` function gives us the actual numbers behind the graphical summaries we just plotted.

```{r}
table(diabetes$frame)
table(diabetes$diabetic, diabetes$frame)
```

Fisher's Exact Test (`fisher.test`) or a chi-squared test (`chiseq.test`) can be used to show that two factors are not independent.

```{r}
fisher.test( table(diabetes$diabetic, diabetes$frame) )
```

> ### Challenge - gender and diabetes {.challenge}
>
> Do you think any association between gender and whether a person is diabetic is shown by this data set?
>
> Why?
>

## Missing data

`summary` gives an overview of a data frame.

```{r}
summary(diabetes)
```

We see that some columns contain `NA`s. `NA` is R's way of indicating missing data. Missing data is important in statistics, so R is very careful with its treatment of this. If we try to calculate with an `NA` the result will be `NA`.

```{r}
1 + NA

mean(diabetes$glyhb)
```

Many summary functions, such as `mean`, have a flag to say ignore `NA` values.

```{r}
mean(diabetes$glyhb, na.rm=TRUE)
```


## Summarizing data frames

We were able to summarize the dimensions (rows or columns) of a matrix with `apply`. In a data frame instead of summarizing along different dimensions, we can summarize with respect to different factor columns.

We already saw how to count different levels in a factor with `table`.

We can use summary functions such as `mean` with a function called `tapply`, which works similarly to `apply`.

```{r}
tapply(diabetes$glyhb, diabetes$frame, mean)
```

We obtain NAs because our data contains NAs. We need to tell `mean` to ignore these. Additional arguments to `tapply` are passed to the function given, here `mean`, so we can tell `mean` to ignore NA with

```{r}
tapply(diabetes$glyhb, diabetes$frame, mean, na.rm=TRUE)
```

The result is a vector, with names from the classifying factor.

We can summarize over several factors, in which case they must be given as a list. Two factors produces a matrix. More factors would produce a higher dimensional *array*.

```{r}
tapply(diabetes$glyhb, list(diabetes$frame, diabetes$gender), mean, na.rm=TRUE)
```


## Melting a matrix into a data frame

You may be starting to see that the idea of a matrix and the idea of a data frame with some factor columns are interchangeable. Depending on what we are doing, we may shift between these two representations of the same data.

Modern R usage emphasizes use of data frames over matrices, as data frames are the more flexible representation. Everything we can represent with a matrix we can represent with a data frame, but not vice versa.

`tapply` took us from a data frame to a matrix. We can go the other way, from a matrix to a data frame, with the `melt` function in the package `reshape2`.

```{r}
library(reshape2)

averages <- tapply(diabetes$glyhb, list(diabetes$frame, diabetes$gender), mean, na.rm=TRUE)
melt(averages)

counts <- table(diabetes$frame, diabetes$gender)
melt(counts)
```

Note: The `aggregate` function effectively combines these two steps for you. See also the `ddply` function in package `plyr`, and the `dplyr` package. There are many variations on the basic idea behind `apply`!


## Merging two data frames
