---
title: For loops
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=4, fig.path = "fig/loops-")
```

# For loops

Suppose we want to print each word in a sentence.
One way is to use six `print` statements:

```{r}
best_practice <- c("Let", "the", "computer", "do", "the", "work")
print_words <- function(sentence) {
  print(sentence[1])
  print(sentence[2])
  print(sentence[3])
  print(sentence[4])
  print(sentence[5])
  print(sentence[6])
}

print_words(best_practice)
```

but that's a bad approach for two reasons:

 1. It doesn't scale: if we want to print the elements in a vector that's hundreds long, we'd be better off just typing them in.

 2. It's fragile: if we give it a longer vector, it only prints part of the data, and if we give it a shorter input, it returns `NA` values because we're asking for elements that don't exist!

```{r}
best_practice[-6]
print_words(best_practice[-6])
```

> ### Tip {.callout}
>
> R has has a special variable, `NA`, for designating missing values that are
> Not Available in a data set. See `?NA` for more details.

Here's a better approach:

```{r}
print_words <- function(sentence) {
  for (word in sentence) {
    print(word)
  }
}

print_words(best_practice)
```

This is shorter---certainly shorter than something that prints every character in a hundred-letter string---and more robust as well:

```{r}
print_words(best_practice[-6])
```

The improved version of `print_words` uses a for loop to repeat an operation---in this case, printing---once for each thing in a collection.
The general form of a loop is:

```{r, eval=FALSE}
for (variable in collection) {
  do things with variable
}
```

We can name the loop variable anything we like (with a few restrictions, e.g. the name of the variable cannot start with a digit).
`in` is part of the `for` syntax.
Note that the body of the loop is enclosed in curly braces `{ }`.
For a single-line loop body, as here, the braces aren't needed, but it is good practice to include them as we did.


Here's another loop that repeatedly updates a variable:

```{r}
len <- 0
vowels <- c("a", "e", "i", "o", "u")
for (v in vowels) {
  len <- len + 1
}
# Number of vowels
len
```

It's worth tracing the execution of this little program step by step.
Since there are five elements in the vector `vowels`, the statement inside the loop will be executed five times.
The first time around, `len` is zero (the value assigned to it on line 1) and `v` is `"a"`.
The statement adds 1 to the old value of `len`, producing 1, and updates `len` to refer to that new value.
The next time around, `v` is `"e"` and `len` is 1, so `len` is updated to be 2.
After three more updates, `len` is 5; since there is nothing left in the vector `vowels` for R to process, the loop finishes.

Note that a loop variable is just a variable that's being used to record progress in a loop.
It still exists after the loop is over, and we can re-use variables previously defined as loop variables as well:

```{r}
letter <- "z"
for (letter in c("a", "b", "c")) {
  print(letter)
}
# after the loop, letter is
letter
```

Note also that finding the length of a vector is such a common operation that R actually has a built-in function to do it called `length`:

```{r}
length(vowels)
```

`length` is much faster than any R function we could write ourselves, and much easier to read than a two-line loop; it will also give us the length of many other things that we haven't met yet, so we should always use it when we can (see this [lesson](01-supp-data-structures.html) to learn more about the different ways to store data in R).


> ### Challenge - Using loops {.challenge}
>
> 1. R has a built-in function called `seq` that creates a list of numbers:
>
> ```{r}
> seq(3)
> ```
>
> Using `seq`, write a function that prints the first **N** natural numbers, one per line:
>
> ```{r, echo=-1}
> print_N <- function(N) {
>   nseq <- seq(N)
>   for (num in nseq) {
>     print(num)
>   }
> }
> print_N(3)
> ```
>
> 2. Write a function called `total` that calculates the sum of the values in a vector.
> (R has a built-in function called `sum` that does this for you.
> Please don't use it for this exercise.)
>
> ```{r, echo=-1}
> total <- function(vec) {
>   #calculates the sum of the values in a vector
>   vec_sum <- 0
>   for (num in vec) {
>     vec_sum <- vec_sum + num
>   }
>   return(vec_sum)
> }
> ex_vec <- c(4, 8, 15, 16, 23, 42)
> total(ex_vec)
> ```
>
> 3. Exponentiation is built into R:
>
> ```{r}
> 2^4
> ```
>
> Write a function called `expo` that uses a loop to calculate the same result.
> ```{r, echo=-1}
> expo <- function(base, power) {
>   result <- 1
>   for (i in seq(power)) {
>     result <- result * base
>   }
>   return(result)
> }
> expo(2, 4)
> ```
>
