---
title: ggplot2
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path = "fig/ggplot-")
```

# Plotting with ggplot2

We already saw some of R's built in plotting facilities with the function `plot`. A more recent and much more powerful plotting library is `ggplot2`. This implements ideas from a book called "The Grammar of Graphics". The syntax is a little strange, but there are plenty of examples in the [online documentation](http://docs.ggplot2.org/current/).

If `ggplot2` isn't already installed, we need to install it.

```{r, eval=FALSE}
install.packages("ggplot2")
```

We then need to load it.

```{r, warning=F}
library(ggplot2)
```

Producing a plot with ggplot2, we must give three things:

1. A data frame containing our data.
2. How the columns of the data frame can be translated into positions, colors, sizes, and shapes of graphical elements ("aesthetics").
3. The actual graphical elements to display ("geometric objects").


## Using ggplot2 with a data frame

Let's use the Gapminder data we loaded in the section on for-loops. If you haven't done this section, you can load it all with:

```{r}
gap <- read.csv("r-intro-files/all-gapminder.csv")
```


```{r}
ggplot(gap, aes(x=year, y=lifeExp)) +
    geom_point()
```

The call to `ggplot` and `aes` sets up the basics of how we are going to represent the various columns of the data frame. `aes` defines the "aesthetics", which is how columns of the data frame map to graphical attributes such as x and y position, color, size, etc. We then literally add layers of graphics to this.

Seasoned programmers may notice with some alarm that `aes` does something very odd, since its bare arguments refer to columns of the data frame as though they were variables. To those programmers we comment briefly that R uses lazy evaluation of function parameters to allow any function to potentially behave like a macro, and that this is exactly as dangerous and powerful as it sounds. It's used tastefully, for the most part.

Further aesthetics can be used. Any aesthetic can be either numeric or categorical, an appropriate scale will be used.

```{r}
ggplot(gap, aes(x=year, y=lifeExp, color=continent, size=pop)) +
    geom_point()
```

A wide variety of geoms are available.

```{r}
ggplot(gap, aes(x=factor(year), y=lifeExp)) +
    geom_boxplot()
```

To draw lines, we need to use a "group" aesthetic.

```{r}
ggplot(gap, aes(x=year, y=lifeExp, group=country, color=continent)) +
    geom_line()
```

`geom_smooth` can be used to show trends.

```{r}
ggplot(gap, aes(x=year, y=lifeExp)) +
    geom_point() +
    geom_smooth()
```

Aesthetics can be specified globally in `ggplot`, or within individual geoms. Here, the group is applied only to draw the lines, and color is used to produce multiple trend lines:

```{r}
ggplot(gap, aes(x=year, y=lifeExp)) +
    geom_line(aes(group=country)) +
    geom_smooth(aes(color=continent))
```


### Highlighting subsets

Geoms can be added that use a different data frame, using the `data=` argument.

```{r}
australia <- gap[gap$country == "Australia",]

ggplot(gap, aes(x=year, y=lifeExp, group=country)) +
    geom_line() +
    geom_line(data=australia, color="red", size=2)
```

Notice also that the second `geom_line` has some further arguments controlling its appearance. These *aren't* aesthetics, they're not a mapping of data to appearance, they are direct specification of the appearance. There isn't an associated scale as when color was an aesthetic.


> ### Challenge - Diabetes data {.challenge}
>
> Let's load our diabetes data frame again.
> 
> ```{r}
> diabetes <- read.csv("r-intro-files/diabetes.csv")
> ```
>
> Experiment with different ways of plotting this data frame. 
>


## Using ggplot2 with a matrix

Let's return to our first matrix example.

```{r}
dat <- read.csv(file="r-intro-files/pvc.csv", row.names=1)
mat <- as.matrix(dat)
```

ggplot only works with data frames, so we need to convert this matrix into data frame form, with one measurement in each row. We can convert to this "long" form with the `melt` function in the library `reshape2`.

```{r}
library(reshape2)
long <- melt(mat)
head(long)
colnames(long) <- c("resin","operator","value")
head(long)
```

```{r}
ggplot(long, aes(x=operator, y=value, group=resin, color=resin)) + 
    geom_line()
```

Notice how ggplot2 is able to use either numerical or categorical (factor) data as x and y coordinates.


## Faceting

Faceting lets us quickly produce a collection of small plots. The plots all have the same scales and the eye can easily compare them.

```{r}
ggplot(long, aes(x=operator, y=value)) + 
    geom_point() +
    facet_wrap(~ resin) 
```

```{r}
ggplot(na.omit(diabetes), aes(x=age, y=glyhb, color=location)) +
    geom_point() +
    facet_grid(gender ~ frame)
```


## Saving ggplots

Ggplots can be saved as we talked about earlier, but with one small twist to keep in mind. The act of plotting a ggplot is actually triggered when it is printed. In an interactive session we are automatically printing each value we calculate, but if you are using a for loop, or other R programming constructs, you might need to explcitly `print( )` the plot.

```{r, eval=FALSE}
# Plot created but not shown.
p <- ggplot(long, aes(x=operator, y=value)) + geom_point()

# Only when we try to look at the value p is it shown
p

# Alternatively, we can explicitly print it
print(p)

# To save to a file
ggsave("test.png", p)

# or
png("test.png")
print(p)
dev.off()
```


> ### Challenge - Gapminder data {.challenge}
>
> Think of a way to summarise the Gapminder data with `tapply` and visualize the summarised data.
>
