---
title: ggplot2
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=4, fig.path = "fig/ggplot-")
```

# Plotting with ggplot2

Let's return to our matrix example.

```{r, results="hide"}
dat <- read.csv(file="data/pvc.csv", row.names=1)
mat <- as.matrix(dat)
```

Instead of storing this data in a matrix, we could store each measurement in a row of a data frame. We can convert to this "long" form with the `melt` function in the library `reshape2`.

```{r}
library(reshape2)
long <- melt(mat)
head(long)
colnames(long) <- c("resin","operator","value")
head(long)
````

This long form is often the best form for exploration and statistical testing of data.

We already saw some of R's built in plotting facilities with the function `plot`. A more recent and much more powerful plotting library is `ggplot2`. This implements ideas from a book called "The Grammar of Graphics". The syntax is a little strange, but there are plenty of examples in the [online documentation](http://docs.ggplot2.org/current/).

If `ggplot2` isn't already installed, we need to install it.

```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
```

A simple example:

```{r}
ggplot(long, aes(x=operator, y=value)) + geom_point()
```

The call to `ggplot` sets up the basics of how we are going to represent the various columns of the data frame. We then literally add layers of graphics to this.

```{r}
ggplot(long, aes(x=operator, y=value)) + geom_boxplot() + geom_point()

ggplot(long, aes(x=operator, y=value, group=resin, color=resin)) +
    geom_line() + theme_bw()

ggplot(long, aes(x=operator, y=value)) +
    facet_wrap(~ resin) + geom_point() + theme_bw()
```

ggplots can be save as we talked about above, but with one small twist to keep in mind. The act of plotting a ggplot is actually triggered when it is printed. In an interactive session we are automatically printing each value we calculate, but if you are writing a function you might need to explcitly `print( )` the plot.

```{r, eval=FALSE}
# Plot created but not shown.
p <- ggplot(long, aes(x=operator, y=value)) + geom_point()

# Only when we try to look at the value p is it shown
p

# Alternatively, we can explicitly print it
print(p)

# To save to a file
png("test.png")
print(p)
dev.off()
```

See also the function `ggsave`.

This long form is also ideal for statistical testing. Here is a one-way ANOVA.

```{r}
oneway.test(value ~ resin, data=long)
```

A two-way ANOVA analysis would be better here. This is *well* beyond the scope of this course, but could be achived with something like:

```{r}
null <- lm(value ~ operator, data=long)
alt <- lm(value ~ resin + operator, data=long)
anova(null, alt)
```
