---
title: ggplot2
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path = "fig/ggplot-")
```

# Plotting with ggplot2

We already saw some of R's built in plotting facilities with the function `plot`. A more recent and much more powerful plotting library is `ggplot2`. This implements ideas from a book called "The Grammar of Graphics". The syntax is a little strange, but there are plenty of examples in the [online documentation](http://docs.ggplot2.org/current/).

If `ggplot2` isn't already installed, we need to install it.

```{r, eval=FALSE}
install.packages("ggplot2")
```

We then need to load it.

```{r}
library(ggplot2)
```

Producing a plot with ggplot2, we must give three things:

1. A data frame containing our data.
2. How the columns of the data frame can be translated into positions, colors, sizes, and shapes of graphical elements ("aesthetics").
3. The actual graphical elements to display ("geometric objects").


## Using ggplot2 with a data frame

Let's load up our diabetes data frame again.

```{r}
diabetes <- read.csv("data/intro-r/diabetes.csv")
```

```{r}
ggplot(diabetes, aes(y=glyhb, x=age)) +
    geom_point()
```

The call to `ggplot` sets up the basics of how we are going to represent the various columns of the data frame. We then literally add layers of graphics to this.

Further aesthetics can be added.

```{r}
ggplot(diabetes, aes(y=glyhb, x=age, size=weight, color=gender)) +
    geom_point()
```


## Using ggplot2 with a matrix

Let's return to our first matrix example.

```{r}
dat <- read.csv(file="data/intro-r/pvc.csv", row.names=1)
mat <- as.matrix(dat)
```

ggplot only works with data frames, so we need to convert this matrix into data frame form, with one measurement in each row. We can convert to this "long" form with the `melt` function in the library `reshape2`.

```{r}
library(reshape2)
long <- melt(mat)
head(long)
colnames(long) <- c("resin","operator","value")
head(long)
```

```{r}
ggplot(long, aes(x=operator, y=value)) + geom_point()
```

Notice how ggplot2 is able to use either numerical or categorical (factor) data as x and y coordinates.

```{r}
ggplot(long, aes(x=operator, y=value)) + geom_boxplot() + geom_point()

ggplot(long, aes(x=operator, y=value, group=resin, color=resin)) +
    geom_line() + theme_bw()

ggplot(long, aes(x=operator, y=value)) +
    facet_wrap(~ resin) + geom_point() + theme_bw()
```


## Saving ggplots

ggplots can be saved as we talked about earlier, but with one small twist to keep in mind. The act of plotting a ggplot is actually triggered when it is printed. In an interactive session we are automatically printing each value we calculate, but if you are using a for loop, or other R programming constructs, you might need to explcitly `print( )` the plot.

```{r, eval=FALSE}
# Plot created but not shown.
p <- ggplot(long, aes(x=operator, y=value)) + geom_point()

# Only when we try to look at the value p is it shown
p

# Alternatively, we can explicitly print it
print(p)

# To save to a file
png("test.png")
print(p)
dev.off()
```

See also the function `ggsave`.
